---
title: "06_maftools_signature_explore"
author: "Ming Tang"
date: "December 13, 2017"
output: html_document
---

### convert annovar files to maf files

`cd /rsrch2/genomic_med/krai/mixing_histology_snakemake/11recount_annovar`

```{r}
library(maftools)
annovar.files<- as.list(dir(".", pattern= "*annovar.txt"))


## some files are empty, so annovarToMaf will give error,
## use possibly http://www.brodrigues.co/blog/2017-03-24-lesser_known_purrr/

x<- function(f){
        ofile<- gsub("_recount_annovar.txt", "", f)
        annovarToMaf(annovar = f, Center = NULL, refBuild = 'hg19', 
                               tsbCol = "sample", basename = ofile)
}

purrr::map(annovar.files, possibly(x, otherwise = NA))



```

merge all the maf files

```{bash}
cd /rsrch2/genomic_med/krai/mixing_histology_snakemake/11recount_annovar
awk 'FNR==1 && NR!=1{next;}{print}' *maf > mixing.maf

rsync -avhP railab:mixing_histology_snakemake/11recount_annovar/mixing.maf data/
```

clean the sample name a bit.

`less -S mixing.maf | sed -E 's/_vs_Pa[0-9]{2}_N//' > mixing_clean.maf`


```{r}
library(maftools)
mixing.maf<- read.maf(here("data/mixing_clean.maf"))

getGeneSummary(mixing.maf)

samples<- getSampleSummary(mixing.maf)
samples$Tumor_Sample_Barcode

samples_not_include<- c("Pa25_T1","Pa25_T2", "Pa28_T1", "Pa28_T2", "Pa29_T3",
                        "Pa32_T1", "Pa32_T2")
samples_to_keep<- setdiff(samples$Tumor_Sample_Barcode, samples_not_include)
mixing.maf<- subsetMaf(mixing.maf, tsb= samples_to_keep )
plotmafSummary(maf = mixing.maf, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE, titvRaw = FALSE)
```


### mutation signature

```{r}

mixing.tnm <- trinucleotideMatrix(maf = mixing.maf, ref_genome = '~/annotations/human/hg19_UCSC_genome/UCSC_hg19_genome.fa', 
                               prefix = 'chr', add = TRUE, useSyn = TRUE)

# on HPC
mixing.tnm <- trinucleotideMatrix(maf = mixing.maf, ref_genome = '/scratch/genomic_med/apps/annot/fasta/hg19_UCSC/UCSC_hg19_genome.fa', 
                               prefix = 'chr', add = TRUE, useSyn = TRUE)

library('NMF')
mixing.sign<- extractSignatures(mat = mixing.tnm, nTry = 6, plotBestFitRes = FALSE)
plotSignatures(mixing.sign, contributions = T)


oncoplot(maf = mixing.maf, showTumorSampleBarcodes =T, top = 20,
drawRowBar = T, drawColBar = T)

### cancer genes
library(here)
cancer_genes<- read_tsv(here("data/cancer_genes/cancer_gene_kel.txt"), col_names = T)

oncoplot(maf = mixing.maf, showTumorSampleBarcodes = T, top = 100,drawRowBar = T, drawColBar = T, removeNonMutated = F, writeMatrix = T, genes= cancer_genes$symbol)

oncoplot(maf = mixing.maf, showTumorSampleBarcodes = T, top = 10,drawRowBar = T, drawColBar = T, removeNonMutated = F, writeMatrix = T, genes= c("TP53", "RB1", "SYNE1", "KMT2C", "KMT2D", "AR", "KDM5C", "KDR", "PDGFRA","PML", "WT1"), sortByAnnotation = TRUE, annotationDat = mixing_meta, clinicalFeatures = c("patient", "smoking", "gender", "response") )

mixing_meta<- read_tsv("data/mixing_meta.txt", col_names = T)


oncoplot(maf = mixing.maf, showTumorSampleBarcodes = T, top = 10, genes = c("KMT2C", "KMT2A", "KMT2D", "TP53", "RB1"),drawRowBar = T, drawColBar = T, removeNonMutated = F, annotationDat = mixing_meta, clinicalFeatures = c("patient", "smoking", "gender", "response"), 
annotationOrder = c("Pa01", "Pa02", "Pa03", "Pa04", "Pa05", "Pa06", "Pa07"))

```

### cancer gene mutations

```{r}

oncoplot(maf = mixing.maf, showTumorSampleBarcodes = T, top = 1000, drawRowBar = T, drawColBar = T, removeNonMutated = F, writeMatrix = T)

sample_order<- mixing.maf@data$Tumor_Sample_Barcode %>% sort() %>% unique() %>% as.character()

oncoplot(maf = mixing.maf, showTumorSampleBarcodes = T, top = 10, drawRowBar = T, drawColBar = T, removeNonMutated = F, writeMatrix = F, sampleOrder = sample_order)

cancer_gene_mut_df<- mixing.maf@data %>% 
        inner_join(cancer_gene_mutations, by = c("Chromosome" = "chromosome", "Start_Position" = "start",
                                                 "End_Position" = "stop", "Reference_Allele" = "read",
                                                 "Tumor_Seq_Allele2" = "variant")) %>%
        dplyr::select(Hugo_Symbol, Tumor_Sample_Barcode, Variant_Classification, amino_acid)
                   
 
cancer_gene_mut_df<- cancer_gene_mut_df %>% 
        distinct() %>% 
        mutate_if(is.factor, as.character) %>% 
        spread(Tumor_Sample_Barcode, Variant_Classification, fill = "")

colnames(cancer_gene_mut_df) %>% unique()
setdiff(metadata$samples, colnames(cancer_gene_mut_df))


cols<- brewer.pal(5, "Dark2")
col.assign<- setNames(cols, unique(metadata_sorted$cancer_type))
rownames(metadata_sorted)<- metadata_sorted$samples
df1<- metadata_sorted[colnames(cancer_gene_mut_mat), "cancer_type", drop = FALSE]


mix.ha1<- HeatmapAnnotation(df = df1, col = list(cancer_type = col.assign))


alter_fun1 = list(
    background = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), gp = gpar(fill = "#CCCCCC", col = NA))
    },
    Missense_Mutation = function(x, y, w, h) {
        grid.rect(x, y, w-unit(0.5, "mm"), h-unit(0.5, "mm"), gp = gpar(fill = "#008000", col = NA))
    }
)

## color for barplot

col_bar1 = c("Missense_Mutation" = "#008000")

mix.cancer.patient.hp1<- oncoPrint(cancer_gene_mut_mat, get_type = function(x) strsplit(x, ";")[[1]],
    alter_fun = alter_fun1, col = col_bar1, 
    column_title = "OncoPrint for mutated cancer genes",
    heatmap_legend_param = list(title = "Alternations", at = "Missense_Mutation", 
        labels = "Missense mutation",
        nrow = 1, title_position = "leftcenter"),
    show_column_names = TRUE, column_order = rev(metadata$samples),
    row_names_gp = gpar(fontsize = 10),
    pct_gp = gpar(fontsize = 10),
    bottom_annotation = mix.ha1)

draw(mix.cancer.patient.hp1, heatmap_legend_side = "bottom", annotation_legend_side = "bottom")

```


### mutational pattern 


convert maf files to vcf files.

```{bash}
cd /rsrch2/genomic_med/krai/mixing_histology_snakemake/11recount_annovar

find *N.maf | parallel 'perl ~/vcf2maf/maf2vcf.pl --input-maf {} --output-dir . --ref-fa /scratch/genomic_med/apps/annot/bundle/human_g1k_v37_decoy.fasta'

cd Users/mtang1/projects/mixing_histology_lung_cancer/data/vcfs

rsync -avhP railab:mixing_histology_snakemake/11recount_annovar/ .
```

```{r}
library(MutationalPatterns)
ref_genome<- "BSgenome.Hsapiens.UCSC.hg19"
library(ref_genome, character.only = TRUE)

vcf_files<- list.files(here("data/vcfs"), pattern = "*.vcf", full.names = TRUE)
basename(vcf_files)
sample_names<- gsub("_vs_.+.vcf", "", basename(vcf_files))
sample_names<- sample_names %>% str_replace("_", "")
vcfs<- read_vcfs_as_granges(vcf_files, sample_names, genome = ref_genome)

## change the chromosome names from 1 to chr1 etc.
# http://kasperdanielhansen.github.io/genbioconductor/html/GenomicRanges_seqinfo.html

#vcfs <- lapply(vcfs, function(x) keepSeqlevels(x, seqlevels(BSgenome.Hsapiens.UCSC.hg19)))

#vcfs <- lapply(vcfs, function(x) seqlengths(x)<- seqlengths(BSgenome.Hsapiens.UCSC.hg19))


mut_mat<- mut_matrix(vcf_list = vcfs, ref_genome = ref_genome)
plot_96_profile(mut_mat[,c(1:19)], ymax = 0.1) 
plot_96_profile(mut_mat[,c(11:20)], ymax = 0.1)

pdf("results/mutation_profile.pdf", 10, 12)
plot_96_profile(mut_mat)
dev.off()

pdf("results/Fig2A.pdf", 10, 12)
plot_96_profile(mut_mat)
dev.off()

#install.packages("lsa")
library(lsa)
cosine(mut_mat[,1], mut_mat[,2])

pdf("results/mutation_spectrum_heatmap.pdf", 8, 6)
Heatmap(cosine(mut_mat), cluster_columns = F, cluster_rows = F, name = "cosine distance\nfor mutation spectrum", top_annotation = mix_ha, cell_fun = cell_fun,
        rect_gp = gpar(type = "none") )
dev.off()
library(corrplot)

corrplot(cosine(mut_mat), method="circle", type="lower")
```

### fit 96 mutation profiles to known mutations signatures

```{r}
sp_url <- paste("http://cancer.sanger.ac.uk/cancergenome/assets/","signatures_probabilities.txt", sep = "")

sp_url <- paste("https://cancer.sanger.ac.uk/cancergenome/assets/", "signatures_probabilities.txt", sep = "")
cancer_signatures = read.table(sp_url, sep = "\t", header = TRUE)

#cancer_signatures<- read.table(sp_url, sep = "\t", header = TRUE)

write_tsv(cancer_signatures, here("data/cosmic_cancer_signatures.txt"))

# Reorder (to make the order of the trinucleotide changes the same)
cancer_signatures<- cancer_signatures[order(cancer_signatures[,1]),]
# Only signatures in matrix
cancer_signatures<- as.matrix(cancer_signatures[,4:33])

fit_res <- fit_to_signatures(mut_mat, cancer_signatures)

select <- which(rowSums(fit_res$contribution) > 0)
select<- order(rowSums(fit_res$contribution), decreasing = TRUE)[1:10]
## use top10 for plotting contribution.

plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], 
                  coord_flip = FALSE, mode = "absolute")

pdf("results/Fig2C.pdf", width= 7, height = 5)
plot_contribution(fit_res$contribution[select,], cancer_signatures[,select], 
                  coord_flip = FALSE, mode = "relative") + 
        theme(axis.text=element_text(size=10, color= "black"),
              axis.text.x = element_text(angle = 90, vjust = 0.2,size = 12),
              axis.title=element_text(size=16,face="bold"),
              plot.title = element_text(size=18))
dev.off()
plot_compare_profiles(mut_mat[,1], fit_res$reconstructed[,1],
                      profile_names = c("Original","Reconstructed \n cancer signatures"))

plot_compare_profiles(mut_mat[,2], fit_res$reconstructed[,2],
                      profile_names = c("Original","Reconstructed \n cancer signatures"))




head(fit_res$contribution)

signature_individual<- function(fit_res){
        df<- as.data.frame(fit_res$contribution)
}
df<- as.data.frame(fit_res$contribution)
df$Signature<- rownames(df)

df  %>% gather("sample", "contribution", 1:19) %>% filter(contribution > 0) %>%
        ggplot(aes(x=sample, y = contribution, fill = Signature)) + geom_bar(stat = "identity", position = "fill")


head(fit_res$contribution)
signature_mat<- fit_res$contribution
colSums(signature_mat)

sig_mat_scaled<- t(t(signature_mat)/colSums(signature_mat) * 100)


cell_fun = function(j, i, x, y, width, height, fill) {
	grid.rect(x = x, y = y, width = width * 0.95, height = height*0.95, 
		gp = gpar(col = "grey", fill = fill, lty = 1, lwd = 0.5))
}

library(readr)
mix_meta<- read_tsv(here("data/final_meta.tsv"), col_names = T)
mix_meta<- mix_meta %>% 
        mutate(cancer_type = case_when(
                cancer_type == "large cell" ~ "LCNEC",
                cancer_type == "adeno" ~ "LUAD",
                cancer_type == "squamous cell" ~ "LUSC",
                cancer_type == "small cell" ~ "SCLC",
                cancer_type == "unknown" ~ "NSCLC-NOS"
        ))

mix_meta<- as.data.frame(mix_meta)
library(RColorBrewer)
set.seed(1)
cols<- brewer.pal(5, "Dark2")
cols.patient<- brewer.pal(9, "Paired")

col.assign<- setNames(cols, unique(mix_meta$cancer_type))
patient.assign<- setNames(cols.patient, unique(mix_meta$patient))
rownames(mix_meta)<- mix_meta$ID
df<- mix_meta[colnames(sig_mat_scaled), c("cancer_type", "patient"), drop = FALSE]

library(ComplexHeatmap)
mix_ha<- HeatmapAnnotation(df = df, col = list(cancer_type = col.assign, patient = patient.assign))
pdf("results/Fig2B.pdf", 8, 6)
Heatmap(sig_mat_scaled, cluster_rows = F, cluster_columns = F, 
        rect_gp = gpar(type = "none"), 
        cell_fun = cell_fun,
        name = "% contribution",
        top_annotation = mix_ha)
dev.off()


Heatmap(sig_mat_scaled[indx,], cluster_rows = F, cluster_columns = F, 
        rect_gp = gpar(type = "none"), 
        cell_fun = cell_fun,
        name = "% contribution",
        top_annotation = mix_ha)
```

stacked barplot 
```{r}
library(tibble)

indx<- order(rowSums(sig_mat_scaled), decreasing = TRUE)[1:10]

colSums(sig_mat_scaled[indx,])

pdf("results/Fig2C.pdf", width= 7, height = 5)
sig_mat_scaled[indx,] %>% 
        as.data.frame() %>%
        rownames_to_column(var="signature") %>%
        mutate(signature=factor(signature, levels = c("Signature.1", "Signature.2", "Signature.3",
                                                     "Signature.4", "Signature.6","Signature.7","Signature.13", "Signature.16", "Signature.21", "Signature.24"))) %>%
        gather(key = "sample", value="contribution", -signature) %>%
        ggplot(aes(x=sample, y = contribution, fill = signature)) +
        geom_bar(stat = "identity", position = "stack") +
        theme_bw(base_size = 18) +
        xlab("") + 
        theme(axis.text=element_text(size=10, color= "black"),
              axis.text.x = element_text(angle = 90, vjust = 0.2,size = 12),
              axis.title=element_text(size=16,face="bold"),
              plot.title = element_text(size=18)) 
dev.off()
```

